The Precambrian paleogeography of Laurentia

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits associated with comments from FM

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

@@

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting of the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia, it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. One explanation is that Tonian extensional basin development is associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the Grenvillian orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggested that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of rifting until after ca. 620 Ma north of the New York promontory.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to be the  result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwestern Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}. The ca. 1.60 Ga Racklan-Forward orogeny in northwestern Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwestern Laurentia A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwestern Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continuous with the South Australia, this interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny on that margin. A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup-Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia ca. 1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path \citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tightly positioned against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwestern Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

The separation of Baltica from Laurentia, once in the configuration shown in Figure \ref{fig:Grenville_reconstructions}, is hypothesized to have occurred ca. 600 Ma associated with the opening of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citealp{Cawood2001a}). The cratons would once again be conjoined ca. 430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.

>>>>

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic, and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

@@

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. Throughout this chapter, I will utilize accretionary orogenesis in a broad sense to refer to the tectonic addition of allochthonous terranes such as island arcs and continental ribbons to Laurentia \citep{Stall2020a}. Collisional orogenesis will refer to orogens interpreted to result from the collision of continent-scale blocks associated with ocean closure \citep{Stall2020a}. As discussed in \cite{Staal2020a}, there is ambiguity in this distinction as both orogen types are the result of collision of lithospheric blocks as the result of subduction with the distinction between these end members largely being made on the basis of scale. Nevertheless, the categorization has paleogeographic utility given that following accretionary orogenesis there will be an ocean basin along the margin (albeit further outboard then before) whereas following collisional orogenesis the margin will have become part of a continental interior.

The overall story of the rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia which experienced metamorphic events between ca. 1.76 and 1.67 Ga (\citealp{Strickland2013a}; Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting of the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia, it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. One explanation is that Tonian extensional basin development is associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the Grenvillian orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggested that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 560 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of earlier rifting until after ca. 620 Ma north of the New York promontory \citep{Allen2010a}.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to be the  result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwestern Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}. The ca. 1.60 Ga Racklan-Forward orogeny in northwestern Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. The U-Pb-Hf signatures of detrital zircons from ca. 1.45 to 1.30 Ga sedimentary rocks of Tasmania have also been interpreted to indicate a Laurentia source consistent with such a paleogeographic connection \citep{Mulder2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwestern Laurentia A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwestern Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continuous with the South Australia, this interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny on that margin. A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup-Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia ca. 1.2 Ga as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path \citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tightly positioned against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwestern Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

Magmatism ca. 615 Ma in both Laurentia and Baltica (in the relative configuration shown in Figure \ref{fig:Grenville_reconstructions}) is interpreted to be associated with the plume-related Central Iapetus Magmatic Province (CIMP; \citealp{Tegner2019a}). This ca. 615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citealp{Cawood2001a, Tegner2019a}). The cratons would once again be conjoined ca. 430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.
